---
title: "Practical Machine Learning Project"
author: "refreform"
date: "`r Sys.Date()`"
output: html_document
---
## Overview/ Data Import  
This project looks at the pml barbell lifts dataset. I divide the data into a 
training and validation set (given the test set is provided as part of the 
assignment, but does not come with the answers).
```{r import, echo=FALSE}
#import data and change variable being predicted into a factor
library(caret)
test<-read.csv("~/pml-testing.csv")
training<-read.csv("~/pml-training.csv")
training$classe<-factor(training$classe)

# Break up training into training and testing
inValidate <- createDataPartition(training$classe, p = 3/4, list = FALSE)
tru_train<-training[-inValidate,]
validation<-training[inValidate,]
```

## Cleaning  
I also clean up the data as there are many columns with almost all null values,
and I remove a couple columns that simply confuse the model, X and user_name. I
also realized all the valuable data was of the type numeric, due to how the 
monitors report data, so I only looked at columns with that datatype. All the 
other columns appeared to simply be noise.
```{r cleaning, echo=FALSE}
#looking at nulls
na_counts <- colSums(is.na(tru_train))
na_prop <- colMeans(is.na(tru_train))
bad_cols <- na_prop > 0.9

#cleaning data by removing columns that are majority NA
tru_train_clean <- tru_train[, !bad_cols]
tru_train_clean<-tru_train_clean[3:93]
validation_clean <- validation[,!bad_cols]
validation_clean<-validation_clean[3:93]

# using only columns that are numerics, as after investigating the data
# it would appear most other columns are either not relevant to predicting
# (such as X and name) or consistently blank
num_cols <- sapply(tru_train_clean, is.numeric)
number_train_clean<-tru_train_clean[,num_cols]
validation_clean<-validation_clean[,num_cols]
number_train_clean<-data.frame(tru_train$classe,number_train_clean)
names(number_train_clean)[1]<-'classe'
validation_clean<-data.frame(validation$classe,validation_clean)
names(validation_clean)[1]<-'classe'
```

## Training  
After cleaning, I train the model on teh training set, which takes quite a 
while. I then look at the prediction set for both the test and validation set.
```{r training, echo=FALSE}
model<-caret::train(classe~.,data=number_train_clean,method='gbm',verbose=FALSE)
```

## Prediction  
My training set has a near 100% accuracy, while the validation set has a
slightly lower accuracy, but not by much. It's still at roughly 99% accuracy. 
This leads me to believe the out of sample error rate is small, <1% in all 
likelihood. 

```{r prediction, echo=FALSE}
prediction_train<-predict(model,newdata=number_train_clean)
confusionMatrix(prediction_train,number_train_clean$classe)

prediction_validate<-predict(model,newdata=validation_clean)
confusionMatrix(prediction_validate,validation_clean$classe)

prediction_test<-predict(model,newdata=test)
```

## Test Set Results  
Lastly, I print the predictions for the 20 test datapoints as requested. 

I have since running this verified with the quiz and had a 95% accuracy with 
those 20 data points.
```{r test restults, echo=FALSE}
prediction_test
```
